---
title: "Are there differences in tap water characteristics across regions in South Australia?"
subtitle: "Assignment 4 ETC5512"
author: "Waranya Ratirotjanakul (36851256)"
date: today
format: html
bibliography: references.bib
csl: apa.csl
nocite: |
  @readxl
  @tidyverse
  @janitor
  @sf_book
  @sf_article
  @here
  @ggrepe
  @rcolorbrewer
  @gt
editor: visual
---

::: panel-tabset
<!-- Task 1 Tab: Data and Documenting -->

## Data and Documenting

### Research Question
Are there differences in tap water characteristics across regions in South Australia?

To answer this question, the analysis addresses three sub-questions:

1. How do tap water characteristics vary across operational regions in South Australia?
2. Which water quality parameters show the greatest variation between regions?
3. How much variation exists between water supply systems within the same operational region?

I chose this question because I am interested in sustainability and environmental systems, particularly how natural resources are managed across different communities. South Australia was selected as the case study because it is one of Australia's driest states [@dew2026waterdryconditions], making water management especially important. After moving from Thailand to Australia, I noticed that drinking tap water directly is much more common in Australia, which made me curious about whether tap water characteristics differ across regions and local water supply systems.

### Data Sources and Suitability

#### Dataset 1: SA Water Quality Performance Results

The primary dataset is the **SA Water Quality Performance Results dataset** published by **SA Water** through the South Australian Government Open Data Portal [@sawater_water_quality]. It contains drinking water quality information, including water quality parameters, average values, guideline values, and disinfection methods, for SA Water operational regions and water supply systems across metropolitan, regional, and remote communities in South Australia.

This dataset is suitable because it includes both regional and system-level information, allowing comparison of tap water characteristics across South Australia and between individual water supply systems. The reported values are based on SA Water's routine monitoring program and represent rolling averages over the previous 12 months. As the data are collected through established monitoring and quality assurance processes, the dataset provides a reliable basis for regional comparison. 

Australian tap water may contain naturally occurring minerals, treatment chemicals, and trace contaminants depending on the water source and distribution system [@completehomefiltration_tap_water]. **Tap water characteristics** refer to the physical and chemical properties of drinking water that influence its quality, safety, and consumer acceptability. These characteristics can be broadly grouped into mineral composition, aesthetic characteristics (such as taste, appearance, and odour), and water treatment characteristics [@nhmrc_physical_chemical_values].

Australian tap water is regulated under the Australian Drinking Water Guidelines (ADWG), which provide health-based and aesthetic guideline values for drinking water quality parameters [@nhmrc_physical_chemical_values]. To represent the main characteristics of tap water, a set of parameters was selected from each category. Together, these parameters provide a broad view of differences in mineral content, water quality, and treatment practices across South Australia.

| Category | Parameter | Justification |
|------------------------|------------------------|------------------------|
| **Mineral composition** | Calcium | Contributes to water hardness and dissolved mineral content. |
|  | Magnesium | Contributes to water hardness and mineral composition. |
|  | Alkalinity | Indicates the water's ability to resist changes in pH. |
|  | Total Dissolved Solids (TDS) | Measures the overall concentration of dissolved minerals in water. |
| **Taste and appearance** | pH | Influences water chemistry and may affect taste. |
|  | Turbidity | Reflects water clarity and the presence of suspended particles. |
|  | Sodium | May influence the salinity and taste of drinking water. |
|  | Sulphate | May contribute to differences in water taste. |
| **Water treatment** | Fluoride | Indicates the use of fluoride treatment in drinking water supplies. |
|  | Chlorine (free residual) | Reflects the level of disinfectant remaining in the water and may affect taste and odour.

**Data type:** Observational data based on water samples collected through routine drinking water quality monitoring. 

**Dataset size:** 4,872 observations and 11 variables

**Data licence:** Creative Commons Attribution 4.0 International (CC BY 4.0)

**Data status:** Active

**Update frequency:** Monthly

**Date published:** 15 June 2018

**Date updated:** 5 May 2026

**Coverage:** South Australia

**Jurisdiction:** South Australian Government

**Publisher/Agency:** SA Water

**Limitations:**

- Results are reported as rolling 12-month average values and may not capture short-term or seasonal variation.
- Operational regions contain multiple water supply systems, which may mask local differences within a region.
- Exact sampling locations are not provided.
The dataset contains many water quality parameters, so representative variables were selected for detailed analysis.
- Each dataset release summarises the previous 12 months. Analysing long-term trends would require combining multiple historical datasets.
- SA Water operational regions do not align directly with South Australian Government regions, requiring an approximate correspondence table for mapping.

**Ethical and privacy considerations:**

- The dataset contains aggregated water quality measurements and does not include personal or identifiable information; therefore, privacy risks are minimal.

#### Dataset 2: South Australian Government Regions

The second dataset is the **South Australian Government Regions dataset** published by the **South Australian Government** [@sa_gov_regions]. It contains administrative regional boundaries used for government planning, monitoring, and service delivery across South Australia.

This dataset is suitable because it provides geographic boundary information that can be used to visualise regional patterns in water quality and create maps for comparison across South Australia.

**Data type:** Administrative geographic data defining official South Australian Government regional boundaries.

**Dataset format:** Spatial boundary dataset (shapefile)

**Data licence:** Creative Commons Attribution 4.0 International (CC BY 4.0)

**Data status:** Active

**Update frequency:** Daily

**Date published:** 21 December 2015

**Date updated:** 20 June 2025

**Geospatial coverage:** South Australia

**Jurisdiction:** South Australian Government

**Publisher/Agency:** Department for Housing and Urban Development

**Limitations:**

- Government regions do not align directly with SA Water operational regions, requiring an approximate correspondence table for mapping.
- The dataset provides geographic boundaries only and does not contain water quality information.
- Regional boundaries are designed for administrative purposes and may not fully represent water supply service areas.

**Ethical and privacy considerations:**

- The dataset contains geographic boundary information only and does not include personal or identifiable information; therefore, privacy risks are minimal.

### Data Download and Processing

#### Dataset 1: SA Water Quality Performance Results

**Download the data**

The dataset was downloaded from the South Australian Government Open Data Portal using the latest available water quality performance results (April 2026) [@sawater_water_quality; @sawater_water_quality_results_Apr2026].

The raw dataset was imported directly from the download URL and saved as a CSV file for reproducibility.

```{r}
# load required packages
library(readxl)
library(tidyverse)

# url of the April 2026 water quality dataset
url_sawater_quality <- "https://data.sa.gov.au/data/dataset/996ec2ae-d52c-4d7e-be9c-d4dab1c1aa45/resource/9d161800-cfdd-46d7-8236-3e601fdc4a1c/download/water-quality-performance-results-april-2026.xlsx"

# download the dataset 
download.file( 
  url_sawater_quality, 
  destfile = "data/sawater_quality_apr26.xlsx", 
  mode = "wb" )

# import the dataset and remove the final 9 footer rows
sawater_quality_apr26 <- read_excel("raw_data/sawater_quality_apr26.xlsx") |>
  slice(1:(n() - 9))

# save the raw dataset as a CSV file
write_csv(
  sawater_quality_apr26,
  "raw_data/sawater_quality_apr26.csv"
  )
```

**Process the data**

The raw dataset (`sawater_quality_apr26.csv`) was processed to prepare it for analysis. The processing steps included standardising variable names, removing unnecessary variables, converting measurement values to numeric format, and selecting water quality parameters relevant to the research question.

```{r}
# load required packages
library(janitor)

# inspect dataset structure and missing values
glimpse(sawater_quality_apr26) 
colSums(is.na(sawater_quality_apr26))

# standardise column names to snake_case
sawater_quality_apr26_processed <- sawater_quality_apr26 |>
  clean_names()

# clean and prepare variables
sawater_quality_apr26_processed <- sawater_quality_apr26_processed |>
  # remove variables not used in the analysis
  select(
    -end_date, 
    -comment) |>
  
  # remove "<" symbol and convert to numeric
  mutate(
    average_value = average_value |>
      str_remove("<") |>
      as.numeric()
  )
  
# inspect processed dataset 
glimpse(sawater_quality_apr26_processed) 
colSums(is.na(sawater_quality_apr26_processed))
```
Parameters were then selected to represent the three main categories of tap water characteristics identified earlier: **mineral composition**, **taste and appearance characteristics**, and **water treatment characteristics**. These parameters were chosen because they provide a broad representation of drinking water quality and consumer acceptability across South Australia. The resulting processed dataset was saved as `sawater_quality_apr26_processed.csv` for subsequent analysis.

```{r}
# check parameter names
unique(sawater_quality_apr26_processed$parameter)

# selected parameters
selected_params <- c(
  "Calcium",
  "Magnesium",
  "Alkalinity",
  "Total dissolved solids",
  "pH",
  "Turbidity",
  "Sodium",
  "Sulphate",
  "Fluoride",
  "Chlorine (free residual)"
)

# retain selected parameters
sawater_quality_apr26_processed <- sawater_quality_apr26_processed |>
  filter(parameter %in% selected_params)

# save processed dataset as a CSV file
write_csv(
  sawater_quality_apr26_processed,
  "processed_data/sawater_quality_apr26_processed.csv"
  )
```

#### Dataset 2: South Australian Government Regions

**Download the data**

The dataset was downloaded from the South Australian Government Open Data Portal as a shapefile containing official South Australian Government regional boundaries [@sa_gov_regions]. The downloaded ZIP file was extracted into the project data folder (`SAGovtRegions_shp`) before being imported into R.

The extracted shapefile package contains both GDA94 and GDA2020 versions of the regional boundaries. The GDA2020 version (`SAGovtRegions_GDA2020.shp`) was selected because it is the current Australian national geodetic datum and provides the most up-to-date spatial reference system available in the dataset.

```{r}
# load required packages 
library(sf)

# import the South Australian Government Regions shapefile
sagovreg_map <- read_sf(
  here::here("raw_data/SAGovtRegions_shp/SAGovtRegions_GDA2020.shp"))
```

**Process the data**

The `sawater_quality_apr26_processed` dataset uses SA Water operational regions, while the South Australia Government Regions dataset (`sagovreg_map`) uses administrative regional boundaries. Because the two datasets use different geographic classifications, a correspondence table was created to approximately align regions for mapping purposes. 

The matching was based on geographic location, regional characteristics, and SA Water's service network information [@sawater_networks_treatment].

| Water Operation Region | Matched SA Regions | Grouping Criteria |
|------------------------|------------------------|------------------------|
| Metropolitan | Eastern Adelaide, Northern Adelaide, Southern Adelaide, Western Adelaide | Adelaide metropolitan and urban areas |
| Outer Metro | Adelaide Hills, Barossa Light and Lower North, Fleurieu and Kangaroo Island | Outer metropolitan and peri-urban areas |
| Eyre | Eyre and Western | Direct geographic alignment |
| Northern | Yorke and Mid North, Far North | Northern inland regional areas |
| South East | Limestone Coast | South-eastern regional alignment |
| Remote Communities | Murray and Mallee | Remote and sparsely populated areas |

This correspondence was used for exploratory visualisation only and does not represent official SA Water operational boundaries.

The processed spatial dataset was saved as `sa_operational_geom.rds`, which contains South Australian Government regional boundaries aggregated to approximate SA Water operational regions and was used for subsequent mapping and visualisation.

```{r}
# create correspondence table between SA Government regions and SA Water operational regions
region_lookup <- tribble(
  ~region, ~water_region,

  "Eastern Adelaide", "Metropolitan",
  "Northern Adelaide", "Metropolitan",
  "Southern Adelaide", "Metropolitan",
  "Western Adelaide", "Metropolitan",

  "Adelaide Hills", "Outer Metro",
  "Barossa, Light and Lower North", "Outer Metro",
  "Fleurieu and Kangaroo Island", "Outer Metro",

  "Eyre and Western", "Eyre",

  "Yorke and Mid North", "Northern",
  "Far North", "Northern",

  "Limestone Coast", "South East",

  "Murray and Mallee", "Remote Communities"
)

# join correspondence table to regional boundaries
sa_operational <- sagovreg_map |>
  left_join(region_lookup, by = "region")

# merge boundaries into operational regions
sa_operational_geom <- sa_operational |>
  group_by(water_region) |>
  summarise()

# inspect processed dataset
glimpse(sa_operational_geom)

# save processed spatial dataset
saveRDS(
  sa_operational_geom, 
  "processed_data/sa_operational_geom.rds" 
  )
```

### Metadata

Metadata files were prepared for both processed datasets (`sawater_quality_apr26_metadata.txt` and `sa_operational_geom_metadata.txt`) and are available in the `processed_data` folder. The metadata files describe the dataset purpose, data sources, processing steps, variables, spatial and temporal coverage, limitations, and licensing information to support interpretation and reproducibility.

### README File

A `README.txt` file was prepared following the Harvard School README template [@harvard_readme_files]. The file documents the processed datasets, data sources, processing steps, and limitations, and is available in the `processed_data` folder.

### Data Dictionary

A data dictionary (`data_dictionary.csv`) was prepared for both processed datasets and is available in the `processed_data` folder. The data dictionary documents variable names, data types, descriptions, example values, and derivation information to support interpretation and reuse of the data.

### Processed Data Folder Structure

All processed datasets and supporting documentation are stored in the `processed_data` folder. The folder structure is shown below.

```text
processed_data/
│
├── README.txt
├── data_dictionary.csv
│
├── sawater_quality_apr26_processed.csv
├── sawater_quality_apr26_metadata.txt
│
├── sa_operational_geom.rds
└── sa_operational_geom_metadata.txt
```

The folder contains the processed datasets used for analysis, together with supporting documentation including metadata files, a data dictionary, and a README file to facilitate interpretation, reproducibility, and reuse.


<!-- Task 2 Tab: Writing your blogpost -->

## Blog post

### Why investigate tap water?

South Australia is one of Australia's driest states [@dew2026waterdryconditions], making effective water management particularly important. Although Australian tap water is generally considered safe to drink, drinking water characteristics can vary depending on water sources, mineral composition, and treatment processes. In South Australia, differences in taste are often associated with higher mineral concentrations and chlorine used during water treatment, even when water quality meets Australian Drinking Water Guidelines (ADWG) [@metropolitanplumbing_adelaide_tap_water].

After moving from Thailand to Australia, one of the things that surprised me was how common it is to drink water directly from the tap. This raised an interesting question: if people across South Australia rely on different water sources and supply systems, do they actually receive water with similar characteristics?

Using publicly available SA Water monitoring data, this blog investigates whether tap water characteristics differ across South Australia, which water quality parameters vary the most between regions, and whether regional averages mask variation between individual water supply systems.

### Where does the data come from?

This analysis uses the SA Water Quality Performance Results dataset published by SA Water through the South Australian Government Open Data Portal [@sawater_water_quality], which contains drinking water quality measurements for water supply systems across South Australia. The reported values are rolling 12-month averages collected through SA Water's routine monitoring program.

To visualise regional patterns, the South Australian Government Regions dataset [@sa_gov_regions] was also used. Since government regions do not align directly with SA Water operational regions, an approximate correspondence table was created for mapping.

The analysis focuses on selected water quality parameters representing mineral composition, taste and appearance characteristics, and water treatment characteristics. These parameters provide a manageable way to compare tap water characteristics across South Australia.

### What does the data tell us?
#### How do tap water characteristics vary across regions?

To compare regional tap water characteristics, three representative parameters were selected: **Total Dissolved Solids (TDS)** for mineral composition, **pH** for taste and appearance characteristics, and **free residual chlorine** for water treatment characteristics. According to the Australian Drinking Water Guidelines (ADWG), the aesthetic guideline for TDS is 600 mg/L, the acceptable pH range is 6.5–8.5, and chlorine concentrations up to 5 mg/L are considered acceptable in drinking water [@nhmrc_physical_chemical_values].

```{r}
#| include: false

# define the water quality parameters selected for spatial analysis
analysis1_params <- c(
  "Total dissolved solids",
  "pH",
  "Chlorine (free residual)"
)

# keep only observations for the selected parameters
analysis1_data <- sawater_quality_apr26_processed |>
  filter(parameter %in% analysis1_params)

# calculate the average water quality value for each parameter within each operational water region
analysis1_summary <- analysis1_data |>
  group_by(region_name, parameter) |>
  summarise(
    avg_value = mean(average_value, na.rm = TRUE),
    .groups = "drop"
  )

# attach the summarised water quality results to the spatial boundaries of SA Water operational regions
map_data <- sa_operational_geom |>
  left_join(
    analysis1_summary,
    by = c("water_region" = "region_name")
  )
```

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: fig-tds_map
#| fig-cap: "Total Dissolved Solids Across South Australia"

# load packages for non-overlapping map labels and colour palettes
library(ggrepel)
library(RColorBrewer)

# extract Total Dissolved Solids (TDS) observations for mapping
tds_map <- map_data |>
  filter(parameter == "Total dissolved solids")

# compute polygon centroids to determine label positions
coords <- st_centroid(tds_map)

# extract centroid coordinates for use in map annotations
coords_xy <- cbind(
  tds_map,
  st_coordinates(coords)
)

# create a choropleth map showing average TDS levels across SA Water operational regions
ggplot(tds_map) +
  geom_sf(aes(fill = avg_value)) +
  
  # add region labels at centroid locations while avoiding overlap
  geom_text_repel(
    data = coords_xy,
    aes(X, Y, label = water_region),
    color = "black",
    fontface = "bold",
    size = 3
  ) +
  
  # apply a sequential blue colour scale where darker shades represent higher TDS concentrations
  scale_fill_distiller(
  palette = "Blues",
  direction = 1,
  labels = scales::comma
) +
  
  # add map title, axis labels and legend title
  labs(
    title = "Total Dissolved Solids Across South Australia",
    x = "Longitude",
    y = "Latitude",
    fill = "TDS (mg/L)"
  ) +

  theme_minimal()
```
```{r}
#| warning: false
#| message: false
#| echo: false
#| label: fig-ph_map
#| fig-cap: "pH Across South Australia"

# filter pH data
ph_map <- map_data |>
  filter(parameter == "pH")

# create region centroids for label placement
coords <- st_centroid(ph_map)

# extract centroid coordinates
coords_xy <- cbind(
  ph_map,
  st_coordinates(coords)
)

# create pH map by operational region
ggplot(ph_map) +
  geom_sf(aes(fill = avg_value)) +
  
  # add region labels
  geom_text_repel(
    data = coords_xy,
    aes(X, Y, label = water_region),
    color = "black",
    fontface = "bold",
    size = 3
  ) +
  
  # apply colour scale
  scale_fill_distiller(
  palette = "Blues",
  direction = 1,
  labels = scales::comma
  ) +
  
  # add title and axis labels
  labs(
    title = "pH Across South Australia",
    x = "Longitude",
    y = "Latitude",
    fill = "pH"
  ) +

  theme_minimal()
```

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: fig-chlorine_map
#| fig-cap: "Free Residual Chlorine Across South Australia"

# filter free residual chlorine data
chlorine_map <- map_data |>
  filter(parameter == "Chlorine (free residual)")

# create region centroids for label placement
coords <- st_centroid(chlorine_map)

# extract centroid coordinates
coords_xy <- cbind(
  chlorine_map,
  st_coordinates(coords)
)

# create free residual chlorine map by operational region
ggplot(chlorine_map) +
  geom_sf(aes(fill = avg_value)) +
  
  # add region labels
  geom_text_repel(
    data = coords_xy,
    aes(X, Y, label = water_region),
    color = "black",
    fontface = "bold",
    size = 3
  ) +
  
  # apply colour scale
  scale_fill_distiller(
  palette = "Blues",
  direction = 1,
  labels = scales::comma
  ) +
  
  # add title and axis labels
  labs(
    title = "Free Residual Chlorine Across South Australia",
    x = "Longitude",
    y = "Latitude",
    fill = "Chlorine concentration (mg/L)"
  ) +

  theme_minimal()
```

@fig-tds_map, @fig-ph_map, @fig-chlorine_map show that **all regions recorded values within ADWG guideline ranges**, indicating that drinking water quality remains acceptable across South Australia. However, noticeable regional differences exist. **TDS showed the largest variation**, with the South East and Eyre regions recording higher mineral concentrations than Metropolitan and Outer Metro regions. Higher TDS values indicate greater concentrations of dissolved minerals, which can influence the taste and perceived hardness of tap water [@nhmrc_physical_chemical_values]. In contrast, pH varied only slightly between regions, while free residual chlorine showed moderate variation, reflecting differences in water treatment and distribution systems.

These maps show that drinking water quality remains within ADWG guideline ranges across South Australia, but noticeable regional differences exist. The largest differences are observed for mineral-related characteristics, particularly TDS, whereas pH remains relatively consistent and chlorine shows only moderate variation. This raises a further question: which water quality parameters contribute most to these regional differences?

#### Which water quality parameters vary the most?
To identify which water quality characteristics contribute most to the observed regional differences, the **coefficient of variation (CV)** was calculated for all selected parameters. Unlike standard deviation, the CV measures variation relative to the average value of each parameter, allowing comparison between variables with different units and scales.

```{r}
#| echo: false
#| label: tbl-summary_stats
#| tbl-cap: "Summary statistics and ADWG guideline values for selected water quality parameters across South Australia"

# summary statistics by parameter
variation_table <- sawater_quality_apr26_processed |>
  group_by(parameter) |>
  summarise(
    mean = mean(average_value, na.rm = TRUE),
    sd = sd(average_value, na.rm = TRUE),
    cv = sd / mean,
    
    # retain guideline values
    health_guideline =
      first(na.omit(health_guideline)),
    
    aesthetic_guideline =
      first(na.omit(aesthetic_guideline)),
    
    .groups = "drop"
  ) |>
  
  # round numeric values
  mutate(
    mean = round(mean, 2),
    sd = round(sd, 2),
    cv = round(cv, 2)
  ) |>
  
  arrange(desc(cv))

library(gt)

variation_table |>
  gt() |>
  cols_label(
    parameter = "Parameter",
    mean = "Mean",
    sd = "SD",
    cv = "CV",
    health_guideline = "Health Guideline",
    aesthetic_guideline = "Aesthetic Guideline"
  ) |>
  fmt_number(
    columns = c(mean, sd, cv),
    decimals = 2
  ) |>
  tab_source_note(
    source_note =
      md("*Note:* Health guideline and aesthetic guideline values are based on the Australian Drinking Water Guidelines (ADWG) 2011.")
  )
```

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: fig-variation_summary
#| fig-cap: "Variation in Water Quality Parameters Across Regions"

ggplot(
  variation_table,
  aes(
    x = reorder(parameter, cv),
    y = cv,
    fill = cv
  )
) +
  geom_col() +
  coord_flip() +
  scale_fill_distiller(
    palette = "Blues",
    direction = 1,
    guide = "none"
  ) +
  labs(
    x = "Parameter",
    y = "Coefficient of Variation"
  ) +
  theme_minimal()
```

@fig-variation_summary and @tbl-summary_stats show that turbidity exhibited the greatest regional variation, followed by several mineral-related parameters including magnesium, calcium, alkalinity, sodium, and sulphate. Although TDS showed clear regional differences in @fig-tds_map, several individual mineral parameters varied even more relative to their average values. In contrast, pH showed very little variation across regions, consistent with @fig-ph_map, while free residual chlorine in @fig-chlorine_map displayed only moderate variation.

Together, these results suggest that regional differences in drinking water are driven primarily by mineral-related characteristics rather than treatment-related characteristics. However, the analysis so far is based on regional averages, which may conceal important differences between individual water supply systems. The next section investigates variation within a single region.

#### Are regional averages hiding local differences?

Because TDS showed clear regional differences in @fig-tds_map and is closely related to dissolved mineral content, it was selected for further investigation. The **Remote Communities region was chosen** because it contains many individual water supply systems and displayed noticeable variation in TDS, allowing local differences to be explored in more detail.

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: fig-remote_tds
#| fig-cap: "TDS Across Water Supply Systems in the Remote Communities Region"

# filter Remote Communities TDS data
remote_tds <- sawater_quality_apr26_processed |>
  filter(
    region_name == "Remote Communities",
    parameter == "Total dissolved solids"
  )

# calculate regional average
remote_avg <- mean(remote_tds$average_value)

ggplot(
  remote_tds,
  aes(
    x = reorder(system_name, average_value),
    y = average_value
  )
) +
  geom_point(
    color = "#2171B5",
    size = 2
  ) +

  # regional average
  geom_hline(
    aes(
      yintercept = remote_avg,
      linetype = "Remote Communities regional average"
    ),
    color = "red",
    linewidth = 0.6
  ) +

  # ADWG aesthetic guideline for TDS
  geom_hline(
    aes(
      yintercept = 600,
      linetype = "ADWG aesthetic guideline"
    ),
    color = "darkred",
    linewidth = 0.6
  ) +

  coord_flip() +

  scale_linetype_manual(
    values = c(
      "Remote Communities regional average" = "dashed",
      "ADWG aesthetic guideline" = "dotted"
    ),
    name = NULL
  ) +

  labs(
    x = "Water Supply System",
    y = "TDS (mg/L)"
  ) +

  theme_minimal()
```

@fig-remote_tds reveals substantial variation between water supply systems. Although the Remote Communities regional average is approximately 360 mg/L, individual systems range from around 30 mg/L in Oak Valley to over 800 mg/L in Watinuma. Several systems exceed the ADWG aesthetic guideline of 600 mg/L, while others remain well below both the guideline and regional average.

This variation suggests that local water supply systems can differ considerably even within the same operational region, likely reflecting differences in source water and mineral composition. Higher TDS values indicate greater concentrations of dissolved minerals, which may contribute to a stronger mineral taste [@metropolitanplumbing_adelaide_tap_water]. Therefore, regional averages are useful for identifying broad patterns but may not fully represent the experience of individual communities.

### So, does tap water differ across South Australia?

This analysis shows that tap water characteristics do vary across South Australia, but the differences are primarily related to mineral composition rather than drinking water safety. While pH remained relatively consistent across regions, larger differences were observed for mineral-related characteristics such as Total Dissolved Solids (TDS), calcium, magnesium, and sulphate.

These differences are likely influenced by the diverse water sources used across South Australia, including River Murray water, reservoirs, groundwater, and desalinated water [@epa_where_water_comes_from; @dew_water_resources_environments]. However, all regions generally remained within Australian Drinking Water Guidelines (ADWG) ranges [@nhmrc_physical_chemical_values].

The analysis also revealed that regional averages can hide substantial local variation. Within the Remote Communities region, TDS values ranged from around 30 mg/L to over 800 mg/L, demonstrating that water characteristics may differ considerably between individual supply systems.

Overall, the results suggest that perceptions of differences in South Australian tap water are not entirely unfounded. While drinking water generally remains within ADWG guideline ranges, differences in mineral composition and source water characteristics can influence taste and consumer perception [@tait2025adelaide_tap_water].


### What could be explored further?

This analysis focused on regional differences in selected water quality characteristics using a single reporting period. Future analyses could explore:

- How tap water characteristics change over time by combining multiple historical SA Water datasets.
- Whether differences in water characteristics can be linked to specific water sources, such as River Murray water, groundwater, reservoirs, or desalinated water.
- Additional water quality parameters not included in this study, such as hardness, aluminium, or nitrate, to provide a more comprehensive assessment of regional water quality.
- The relationship between measured water quality characteristics and consumer perceptions of taste and drinking water quality.


<!-- Task 3 Tab: Behind the Scenes -  -->

## Behind the Scenes

### Question 5
*Tell us about parts of your data processing or analysis that weren’t “sexy” and wouldn’t typically be included in a blog post. (e.g. Was their any data drudgery or time intensive wrangling? Were there any repetitive tasks or manual tasks? If it was easy, describe what made it easy?)*

Some of the less visible but important tasks in this project included:

- I spent time cleaning the SA Water dataset by removing footer rows, standardising variable names, and converting values reported below detection limits into numeric format.
- I reviewed a large number of water quality parameters before selecting a smaller set that best represented mineral composition, taste and appearance, and water treatment characteristics.
- I manually created a correspondence table between SA Water operational regions and South Australian Government regions so that the water quality data could be mapped spatially.
- I prepared metadata, data dictionaries, and a README file to ensure the processed datasets were well documented and reproducible.
- I tested a range of visualisation approaches before deciding on the final set of maps and figures used in the blog post.


### Question 6
*Were there any challenges that you faced in conducting this analysis. These may take the form of data limitations or coding challenges? (e.g. Was there anything in your analysis that you were not anticipating when you started? Did you have to change your intended scope? Did you need to master a new skill? Were there any problems you were proud of solving?)*

Some challenges encountered during this project included:

- The SA Water dataset uses operational regions, while the spatial dataset uses South Australian Government regions. Because no official correspondence table was available, I had to develop an approximate mapping manually.
- I initially attempted to display all selected water quality parameters using faceted maps. However, large differences in units and value ranges made comparisons difficult and sometimes visually misleading, so I refined the scope to focus on three representative parameters.
- The dataset does not contain detailed information about water sources for individual supply systems, limiting my ability to explain the exact causes of regional differences.
- I originally considered analysing changes in water quality over time. However, combining multiple historical SA Water datasets was beyond the scope of this project.

A key skill developed during this project was working with spatial data in R, including joining spatial and tabular datasets, creating centroids for maps labels, and producing maps to investigate regional patterns.

I was particularly proud of solving the mismatch between the two regional classification systems and creating operational region maps that could be linked to the water quality data. This required manually developing a correspondence table because no official mapping was available, and it enabled the regional visualisations that formed a key part of the analysis.


### Question 7
*Tell us about any imperfect parts of your work and how you would like to expand or improve this analysis in future? Be clear about any limitations or aspects of your analysis that fell beyond scope.*

Some limitations of this analysis and potential future improvements include:

- I used data from a single reporting period, which provides only a snapshot of tap water characteristics. In future, I would like to combine multiple historical SA Water datasets to investigate how water quality changes over time.
- I focused on a selected set of water quality parameters to keep the analysis manageable. Including additional parameters could provide a more comprehensive assessment of regional water quality.
- Although the analysis identified regional differences in mineral-related characteristics, the dataset does not contain detailed information about water sources for individual supply systems. In future, I would like to investigate whether these differences are associated with specific water sources such as River Murray water, groundwater, reservoirs, or desalinated water.
- This study focused on measured water quality characteristics rather than consumer experiences. Future work could explore whether differences in water quality parameters are related to public perceptions of tap water taste and quality.


### Question 8
*Tell us about how you used AI in your approach to your assignment. (e.g. Was AI consistently reliable? Did you provide AI with additional context? Was there anywhere where you questioned or verified the AI outputs? Was there anywhere we AI helped you learn something new?)*

AI was used as a support tool throughout this project for brainstorming research questions, identifying meaningful analyses within the limitations of the available dataset, assisting with R coding, and improving documentation and written communication. It helped me explore analytical approaches, such as using the coefficient of variation (CV) to compare variables with different units and scales, and learn new techniques such as creating centroids and adding labels to spatial maps in R. To obtain relevant responses, I provided detailed context about the datasets, research objectives, and assignment requirements. However, AI outputs were not accepted without review. I verified coding suggestions, interpretations, and conclusions against the data, relevant literature, and assignment requirements before incorporating them into the final submission.


### Question 9
*Submit 4 earlier versions of your .qmd files to show your iterative process. These versions must show iterations of your work (not just a linear progression of Task 1, Task 2, Task 3, citations). Reflect on the progress made between each versions by providing a short statement for markers of what you fixed/learnt/improved/changed between each file. Also like to reflect on what changed between your first checklist submission. (Note: There is an expectation for students familiar with GitHub that you will submit a Github repository and refer to 4 individual commits.)*

| Commit | Commit message | Iterations of work
|--------|----------------|-------------------------
|        | Refine reflection section and clarify limitations and future work | Refined the reflection section by reducing overlap between Questions 5–7. I separated behind-the-scenes work, analytical challenges, and project limitations into distinct sections, while also clarifying future directions for extending the analysis. This improved the overall consistency between the blog post, conclusions, and reflections.
|.       |

:::
