---
title: "What's in Your Tap Water? Exploring Regional Differences Across South Australia"
subtitle: "Assignment 4 ETC5512"
author: "Waranya Ratirotjanakul (36851256)"
date: today
format: html
bibliography: references.bib
csl: apa.csl
nocite: |
  @readxl
  @tidyverse
  @janitor
  @sf_book
  @sf_article
  @here
editor: visual
---

::: panel-tabset
<!-- Task 1 Tab: Data and Documenting -->

## Data and Documenting

### Research Question
Are there differences in tap water characteristics across regions in South Australia?

To answer this question, the analysis addresses three sub-questions:

1. How do tap water characteristics vary across operational regions in South Australia?
2. Which water quality parameters show the greatest variation between regions?
3. How much variation exists between water supply systems within the same operational region?

I chose this question because I am interested in sustainability and environmental systems, particularly how natural resources are managed across different communities. South Australia was selected as the case study because it is one of Australia's driest states [@dew2026waterdryconditions], making water management especially important. After moving from Thailand to Australia, I noticed that drinking tap water directly is much more common in Australia, which made me curious about whether tap water characteristics differ across regions and local water supply systems.

### Data Sources and Suitability

#### Dataset 1: SA Water Quality Performance Results

The primary dataset is the **SA Water Quality Performance Results dataset** published by **SA Water** through the South Australian Government Open Data Portal [@sawater_water_quality]. It contains drinking water quality information, including water quality parameters, average values, guideline values, and disinfection methods, for SA Water operational regions and water supply systems across metropolitan, regional, and remote communities in South Australia.

This dataset is suitable because it includes both regional and system-level information, allowing comparison of tap water characteristics across South Australia and between individual water supply systems. The reported values are based on SA Water's routine monitoring program and represent rolling averages over the previous 12 months. As the data are collected through established monitoring and quality assurance processes, the dataset provides a reliable basis for regional comparison. 

Australian tap water may contain naturally occurring minerals, treatment chemicals, and trace contaminants depending on the water source and distribution system [@completehomefiltration_tap_water]. **Tap water characteristics** refer to the physical and chemical properties of drinking water that influence its quality, safety, and consumer acceptability. These characteristics can be broadly grouped into mineral composition, aesthetic characteristics (such as taste, appearance, and odour), and water treatment characteristics [@nhmrc_physical_chemical_values].

Australian tap water is regulated under the Australian Drinking Water Guidelines (ADWG), which provide health-based and aesthetic guideline values for drinking water quality parameters [@nhmrc_physical_chemical_values]. To represent the main characteristics of tap water, a set of parameters was selected from each category. Together, these parameters provide a broad view of differences in mineral content, water quality, and treatment practices across South Australia.

| Category | Parameter | Justification |
|------------------------|------------------------|------------------------|
| **Mineral composition** | Calcium | Contributes to water hardness and dissolved mineral content. |
|  | Magnesium | Contributes to water hardness and mineral composition. |
|  | Alkalinity | Indicates the water's ability to resist changes in pH. |
|  | Total Dissolved Solids (TDS) | Measures the overall concentration of dissolved minerals in water. |
| **Taste and appearance** | pH | Influences water chemistry and may affect taste. |
|  | Turbidity | Reflects water clarity and the presence of suspended particles. |
|  | Sodium | May influence the salinity and taste of drinking water. |
|  | Sulphate | May contribute to differences in water taste. |
| **Water treatment** | Fluoride | Indicates the use of fluoride treatment in drinking water supplies. |
|  | Chlorine (free residual) | Reflects the level of disinfectant remaining in the water and may affect taste and odour.

**Data type:** Observational data based on water samples collected through routine drinking water quality monitoring. 

**Dataset size:** 4,872 observations and 11 variables

**Data licence:** Creative Commons Attribution 4.0 International (CC BY 4.0)

**Data status:** Active

**Update frequency:** Monthly

**Date published:** 15 June 2018

**Date updated:** 5 May 2026

**Coverage:** South Australia

**Jurisdiction:** South Australian Government

**Publisher/Agency:** SA Water

**Limitations:**

- Results are reported as rolling 12-month average values and may not capture short-term or seasonal variation.
- Operational regions contain multiple water supply systems, which may mask local differences within a region.
- Exact sampling locations are not provided.
The dataset contains many water quality parameters, so representative variables were selected for detailed analysis.
- Each dataset release summarises the previous 12 months. Analysing long-term trends would require combining multiple historical datasets.
- SA Water operational regions do not align directly with South Australian Government regions, requiring an approximate correspondence table for mapping.

**Ethical and privacy considerations:**

- The dataset contains aggregated water quality measurements and does not include personal or identifiable information; therefore, privacy risks are minimal.

#### Dataset 2: South Australian Government Regions

The second dataset is the **South Australian Government Regions dataset** published by the **South Australian Government** [@sa_gov_regions]. It contains administrative regional boundaries used for government planning, monitoring, and service delivery across South Australia.

This dataset is suitable because it provides geographic boundary information that can be used to visualise regional patterns in water quality and create maps for comparison across South Australia.

**Data type:** Administrative geographic data defining official South Australian Government regional boundaries.

**Dataset format:** Spatial boundary dataset (shapefile)

**Data licence:** Creative Commons Attribution 4.0 International (CC BY 4.0)

**Data status:** Active

**Update frequency:** Daily

**Date published:** 21 December 2015

**Date updated:** 20 June 2025

**Geospatial coverage:** South Australia

**Jurisdiction:** South Australian Government

**Publisher/Agency:** Department for Housing and Urban Development

**Limitations:**

- Government regions do not align directly with SA Water operational regions, requiring an approximate correspondence table for mapping.
- The dataset provides geographic boundaries only and does not contain water quality information.
- Regional boundaries are designed for administrative purposes and may not fully represent water supply service areas.

**Ethical and privacy considerations:**

- The dataset contains geographic boundary information only and does not include personal or identifiable information; therefore, privacy risks are minimal.

### Data Download and Processing

#### Dataset 1: SA Water Quality Performance Results

**Download the data**

The dataset was downloaded from the South Australian Government Open Data Portal using the latest available water quality performance results (April 2026) [@sawater_water_quality; @sawater_water_quality_results_Apr2026].

The raw dataset was imported directly from the download URL and saved as a CSV file for reproducibility.

```{r}
# load required packages
library(readxl)
library(tidyverse)

# url of the April 2026 water quality dataset
url_sawater_quality <- "https://data.sa.gov.au/data/dataset/996ec2ae-d52c-4d7e-be9c-d4dab1c1aa45/resource/9d161800-cfdd-46d7-8236-3e601fdc4a1c/download/water-quality-performance-results-april-2026.xlsx"

# download the dataset 
download.file( 
  url_sawater_quality, 
  destfile = "data/sawater_quality_apr26.xlsx", 
  mode = "wb" )

# import the dataset and remove the final 9 footer rows
sawater_quality_apr26 <- read_excel("raw_data/sawater_quality_apr26.xlsx") |>
  slice(1:(n() - 9))

# save the raw dataset as a CSV file
write_csv(
  sawater_quality_apr26,
  "raw_data/sawater_quality_apr26.csv"
  )
```

**Process the data**

The raw dataset (`sawater_quality_apr26.csv`) was processed to prepare it for analysis. The processing steps included standardising variable names, removing unnecessary variables, converting measurement values to numeric format, and selecting water quality parameters relevant to the research question.

```{r}
# load required packages
library(janitor)

# inspect dataset structure and missing values
glimpse(sawater_quality_apr26) 
colSums(is.na(sawater_quality_apr26))

# standardise column names to snake_case
sawater_quality_apr26_processed <- sawater_quality_apr26 |>
  clean_names()

# clean and prepare variables
sawater_quality_apr26_processed <- sawater_quality_apr26_processed |>
  # remove variables not used in the analysis
  select(
    -end_date, 
    -comment) |>
  
  # remove "<" symbol and convert to numeric
  mutate(
    average_value = average_value |>
      str_remove("<") |>
      as.numeric()
  )
  
# inspect processed dataset 
glimpse(sawater_quality_apr26_processed) 
colSums(is.na(sawater_quality_apr26_processed))
```
Parameters were then selected to represent the three main categories of tap water characteristics identified earlier: **mineral composition**, **taste and appearance characteristics**, and **water treatment characteristics**. These parameters were chosen because they provide a broad representation of drinking water quality and consumer acceptability across South Australia. The resulting processed dataset was saved as `sawater_quality_apr26_processed.csv` for subsequent analysis.

```{r}
# check parameter names
unique(sawater_quality_apr26_processed$parameter)

# selected parameters
selected_params <- c(
  "Calcium",
  "Magnesium",
  "Alkalinity",
  "Total dissolved solids",
  "pH",
  "Turbidity",
  "Sodium",
  "Sulphate",
  "Fluoride",
  "Chlorine (free residual)"
)

# retain selected parameters
sawater_quality_apr26_processed <- sawater_quality_apr26_processed |>
  filter(parameter %in% selected_params)

# save processed dataset as a CSV file
write_csv(
  sawater_quality_apr26_processed,
  "processed_data/sawater_quality_apr26_processed.csv"
  )
```

#### Dataset 2: South Australian Government Regions

**Download the data**

The dataset was downloaded from the South Australian Government Open Data Portal as a shapefile containing official South Australian Government regional boundaries [@sa_gov_regions]. The downloaded ZIP file was extracted into the project data folder (`SAGovtRegions_shp`) before being imported into R.

The extracted shapefile package contains both GDA94 and GDA2020 versions of the regional boundaries. The GDA2020 version (`SAGovtRegions_GDA2020.shp`) was selected because it is the current Australian national geodetic datum and provides the most up-to-date spatial reference system available in the dataset.

```{r}
# load required packages 
library(sf)

# import the South Australian Government Regions shapefile
sagovreg_map <- read_sf(
  here::here("raw_data/SAGovtRegions_shp/SAGovtRegions_GDA2020.shp"))
```

**Process the data**

The `sawater_quality_apr26_processed` dataset uses SA Water operational regions, while the South Australia Government Regions dataset (`sagovreg_map`) uses administrative regional boundaries. Because the two datasets use different geographic classifications, a correspondence table was created to approximately align regions for mapping purposes. 

The matching was based on geographic location, regional characteristics, and SA Water's service network information [@sawater_networks_treatment].

| Water Operation Region | Matched SA Regions | Grouping Criteria |
|------------------------|------------------------|------------------------|
| Metropolitan | Eastern Adelaide, Northern Adelaide, Southern Adelaide, Western Adelaide | Adelaide metropolitan and urban areas |
| Outer Metro | Adelaide Hills, Barossa Light and Lower North, Fleurieu and Kangaroo Island | Outer metropolitan and peri-urban areas |
| Eyre | Eyre and Western | Direct geographic alignment |
| Northern | Yorke and Mid North, Far North | Northern inland regional areas |
| South East | Limestone Coast | South-eastern regional alignment |
| Remote Communities | Murray and Mallee | Remote and sparsely populated areas |

This correspondence was used for exploratory visualisation only and does not represent official SA Water operational boundaries.

The processed spatial dataset was saved as `sa_operational_geom.rds`, which contains South Australian Government regional boundaries aggregated to approximate SA Water operational regions and was used for subsequent mapping and visualisation.

```{r}
# create correspondence table between SA Government regions and SA Water operational regions
region_lookup <- tribble(
  ~region, ~water_region,

  "Eastern Adelaide", "Metropolitan",
  "Northern Adelaide", "Metropolitan",
  "Southern Adelaide", "Metropolitan",
  "Western Adelaide", "Metropolitan",

  "Adelaide Hills", "Outer Metro",
  "Barossa, Light and Lower North", "Outer Metro",
  "Fleurieu and Kangaroo Island", "Outer Metro",

  "Eyre and Western", "Eyre",

  "Yorke and Mid North", "Northern",
  "Far North", "Northern",

  "Limestone Coast", "South East",

  "Murray and Mallee", "Remote Communities"
)

# join correspondence table to regional boundaries
sa_operational <- sagovreg_map |>
  left_join(region_lookup, by = "region")

# merge boundaries into operational regions
sa_operational_geom <- sa_operational |>
  group_by(water_region) |>
  summarise()

# inspect processed dataset
glimpse(sa_operational_geom)

# save processed spatial dataset
saveRDS(
  sa_operational_geom, 
  "processed_data/sa_operational_geom.rds" 
  )
```

### Metadata

Metadata files were prepared for both processed datasets (`sawater_quality_apr26_metadata.txt` and `sa_operational_geom_metadata.txt`) and are available in the `processed_data` folder. The metadata files describe the dataset purpose, data sources, processing steps, variables, spatial and temporal coverage, limitations, and licensing information to support interpretation and reproducibility.

### README File

A `README.txt` file was prepared following the Harvard School README template [@harvard_readme_files]. The file documents the processed datasets, data sources, processing steps, and limitations, and is available in the `processed_data` folder.

### Data Dictionary

A data dictionary (`data_dictionary.csv`) was prepared for both processed datasets and is available in the `processed_data` folder. The data dictionary documents variable names, data types, descriptions, example values, and derivation information to support interpretation and reuse of the data.

### Processed Data Folder Structure

All processed datasets and supporting documentation are stored in the `processed_data` folder. The folder structure is shown below.

```text
processed_data/
│
├── README.txt
├── data_dictionary.csv
│
├── sawater_quality_apr26_processed.csv
├── sawater_quality_apr26_metadata.txt
│
├── sa_operational_geom.rds
└── sa_operational_geom_metadata.txt
```

The folder contains the processed datasets used for analysis, together with supporting documentation including metadata files, a data dictionary, and a README file to facilitate interpretation, reproducibility, and reuse.


<!-- Task 2 Tab: Writing your blogpost -->

## Blog post

### What's in this section

Here is where you should write your blogpost! (Task 2)

### Blogpost Structure

There is no strict structure for you to follow. However, here is a skeleton structure you may find helpful.

1.  Title (is set at the top of this document)
2.  Motivation\
3.  Data\
4.  Analysis\
5.  Conclusions\
6.  References

<!-- Task 3 Tab: Behind the Scenes -  -->

## Behind the Scenes

### What's in this section

Here is where you should tell us about your reflection on your analysis (Task 3).

Again, these are the details about **your** perspective and the gritty details behind the scenes of your analysis.

:::
